---
title: "Evaluation of basketball statistics"
author: "Blake Nelson, Nick Olech"
date: "2024-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description of Data Set

For our project we are combining two different data sets.  The first data frame has information for each player such as personal information and draft information like what position, draft year, and draft team.  The second dataframe gives year by year information for each player in the first data frame. This information includes salary, season, and team.

### Loading packages and data frames
```{r, message=FALSE, error=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(dplyr)
```

```{r}
#Draft information Data Frame
df1 <-
  read.csv(
    "https://query.data.world/s/7fyub35dc6n4crgjosveptbc7uxeo3?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )

#Salary information Data Frame
df2 <-
  read.csv(
    "https://query.data.world/s/6xhwdm7emdeglx4botceotj6udfevo?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )
```

# First Data Cleaning Steps

```{r}
#Merge df1 (draft info) and df2 (salary statistics)
nba_stats <-
  merge(df2,
        df1,
        by.x = "player_id",
        by.y = "X_id",
        all.x = TRUE)
```

```{r}
#Drop player that were drafted before 1980
nba_stats <- nba_stats %>%
  filter(draft_year >= 1980)

#Replaces empty values with NA
nba_stats <- nba_stats %>% 
  mutate(across(where(is.character), ~na_if(.x, "")))
```

# Marginal Summaries
```{r}
head(nba_stats)
```

```{r}
summary(nba_stats)
```

```{r}
dim(nba_stats)
```

# Questions to possibly research

What is the ideal height/weight (build) of an NBA Player?
- Could look at how salary relates to certain measurements a player has.

Does where you are born influence your likely hood of NBA success?
- Could look at salary or sheer number of NBA players from certain regions and then divide by each respective population.

What position in basketball seems to contribute to winning the most?
- Could look to analyze win shares and other statistics relating to positions.

Ideas
Look to model the perfect NBA "prospect"
Utilize comparative analysis in order to predict future player success.

