---
title: "DS Final"
author: "Blake Nelson, Nick Olech"
date: "2024-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Merging 2 data frames

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)

#(PUF2015_021518)
df <-
  read.csv(
    "https://query.data.world/s/7fyub35dc6n4crgjosveptbc7uxeo3?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )


df2 <-
  read.csv(
    "https://query.data.world/s/6xhwdm7emdeglx4botceotj6udfevo?dws=00000",
    header = TRUE,
    stringsAsFactors = FALSE
  )

#Merges df (draft info and general stats) and df2 (salary statistics)
nba_stats <-
  merge(df2,
        df1,
        by.x = "player_id",
        by.y = "X_id",
        all.x = TRUE)

head(nba_stats)
dim(nba_stats)
summary(nba_stats)
```
Questions to possibly research

What is the ideal height/weight (build) of an NBA Player?
- Could look at how salary relates to certain measurements a player has.

Does where you are born influence your likely hood of NBA success?
- Could look at salary or sheer number of NBA players from certain regions and then divide by each respective population.

What position in basketball seems to contribute to winning the most?
- Could look to analyze win shares and other statistics relating to positions.

Ideas
Look to model the perfect NBA "prospect"
Utilize comparative analysis in order to predict future player success.

```{r}
#Gets rid of players that were drafted before 1980
nba_stats <- nba_stats %>%
  filter(draft_year >= 1980)

#Gets rid of empty row values and replaces them with NA
nba_stats <- nba_stats %>% 
  mutate(across(where(is.character), ~na_if(.x, "")))

```

